Fast solvers for the high-order FEM simplicial de Rham complex \thanks{Submitted to the editors July XX, 2024. \funding{ PDB and PEF were supported by EPSRC grant EP/W026260/1. } }

### 2024-06-18 paper/paper.tex
propagate notation

Main challenges, motivations
\begin{itemize}
\item In high-order FEM, the curse of dimensionality affects operator
   evaluation and relaxation methods in multigrid and domain-decomposition.  
\item For tensor product cells,
   a tensor product basis will expose the structure to decouple the
   multidimensional problems into a sequence of one-dimensional problems.
   Sum-factorization enables fast operator evaluation, and the fast
   diagonalization method addresses the efficient solution of problems
   structured subdomains.  On unstructured subdomains, an optimal strategy
   using static condensation with FDM bases was proposed in our previous
   work.
\item For simplices, Bernstein--Bezier expansions enable sum-factorized
   operator evaluation, but there is no knowledge of basis transformation
   that can diagonalize a problem on any structured or unstructured subdomain.
\end{itemize}

@@

For each finite element subcomplex built on a mesh $\mesh$ of a domain $\Omega \subset \mathbb{R}^d$, we denote the spaces discretizing $k$-forms $H\Lambda^k(\Omega)$ with polynomial degree $p$ as $X^k_p(\mesh)$.
These are built from cellwise spaces via the standard pullback construction from a reference cell $\Khat$:
\begin{equation}
X^k_p(\mesh) = \{ v \in H\Lambda^k(\Omega): \Forall K \in \mesh \Exists \hat{v} \in X^k_p(\Khat) \text{ s.t.} \left.v\right|_K = \pullback^k_K(\hat{v}) \},
\end{equation}
where $\pullback^k_K$ denotes the appropriate pullback that preserves continuity of the traces of a $k$-form across cell facets.

@@

Recall the definition of the $\Hcurl$-conforming finite element spaces
\begin{equation}
V^1_{p, p_E}(K) = \{v\in [\P_p(K)]^d : v\cdot\bt \in \P_{p_E}(E) \}
\end{equation}
for $p_E=p-1$ we obtain the \Nedelec space of the first kind $\Ned_p(K)$, while
$p_E=p$ gives the \Nedelec space of the second kind $\NedTwo_p(K) = [\P_p(K)]^d$. 
These spaces enforce continuity of the tangential trace across mesh interfaces.

We define a basis for the dual of $V^1_{p, p_E}(K)$ as 
tangential moments along $E\in \text{edges}(\Khat)$,
\begin{equation}
   \ell^E_j(v) = (q_j, v\cdot \bt)_E, \quad q_j \in \P_{p_E}(E), 
\end{equation}
and for each sub-entity $S \in \text{faces}(\Khat) \cup \text{interior}(\Khat)$,
\begin{align}
   \ell^{S,0}_j(v) &= (\grad_S\phi^S_j, v)_{S}, \\
   \ell^{S,1}_j(v) &= (\curl_S\Phi^S_j, \curl_S v)_{S}, \label{eq:hcurl-interior-dofs}
\end{align}
where $\curl_S$ is the tangential curl on $S$,
$\{\phi^S_j\}$ are the $H_0(\grad, S)$ bases constructed in \eqref{eq:hgrad-fdm},
and $\{\curl_S\Phi^S_j\}$ is a basis for $\curl_S \mathbb{B}_{p_S}(\curl, S)$
such that
\begin{equation} \label{eq:hcurl-fdm}
   (\curl_S\Phi^S_j, \curl_S\Phi^S_i)_{S} = \delta_{ij}, \quad
   (\Phi^S_j, \Phi^S_i)_{S} = \lambda_j\delta_{ij}, \quad
   \Phi^S_j\times\bn = 0 \text{ on }\partial S.
\end{equation}

@@

Recall the definition of the $\Hdiv$-conforming finite element spaces
\begin{equation}
V^2_{p, p_F}(K) = \{v\in [\P_p(K)]^d : v\cdot\bn \in \P_{p_F}(F) \}
\end{equation}
for $p_F=p-1$ we obtain the Raviart--Thomas space $\RT_p(K)$, while
$p_F=p$ gives the Brezzi--Douglas--Marini space $\BDM_p(K) = [\P_p(K)]^d$. 
These spaces enforce continuity of the normal trace across mesh interfaces.

We define a basis for the dual of $V^2_{p, p_F}(K)$ as normal moments on $F\in
\text{faces}(\Khat)$,
\begin{equation}
   \ell^F_j(v) = (q_j, v\cdot \bn)_F, \quad q_j \in \P_{p_F}(F),
\end{equation}
and on the $\text{interior}(\Khat) = K$: 
\begin{align}
   \ell^{K,0}_j(v) &= (\curl\Phi^K_j, v)_{K}, \\
   \ell^{K,1}_j(v) &= (\div\Psi^K_j, \div v)_{K}, \label{eq:hdiv-interior-dofs}
\end{align}
where $\{\Phi^K_j\}$ are the $H_0(\curl, S)$ bases constructed in
\eqref{eq:hcurl-fdm}, and $\{\div\Psi^K_j\}$ is a basis for $\div
\mathbb{B}_p(\div, K)$ such that
\begin{equation} \label{eq:hdiv-fdm}
   (\div\Psi^K_j, \div\Psi^K_i)_K = \delta_{ij}, \quad
   (\Psi^K_j, \Psi^K_i)_{K} = \lambda_j\delta_{ij}, \quad
   \Psi^K_j\cdot\bn = 0 \text{ on }\partial K.
\end{equation}

@@

For 0-forms we construct patches on vertex-stars,
\begin{equation}
   V^0_{h,p} = \left.V^0_{h,p}\right|_\mathcal{I} + V^0_{h,1} 
   + \sum_{v\in E^0(\mesh)} \left.\tilde{V}^0_{h,p}\right|_{\star v} 
\end{equation}

For 1-forms we construct edge-star patches
plus the gradient of vertex-star patches on 0-forms
\begin{equation}
   V^1_{h,p} = \left.V^1_{h,p}\right|_\mathcal{I} +  V^1_{h,1}
   + \sum_{e\in E^1(\mesh)} \left.\tilde{V}^1_{h,p}\right|_{\star e} 
   + \sum_{v\in E^0(\mesh)} \mathrm{d} \left.\tilde{V}^0_{h,p}\right|_{\star v} 
\end{equation}

For 2-forms we construct edge-star patches,
\begin{equation}
   V^2_{h,p} = \left.V^2_{h,p}\right|_\mathcal{I} +  V^2_{h,1} 
   + \sum_{e\in E^1(\mesh)} \left.\tilde{V}^2_{h,p}\right|_{\star e}. 
\end{equation}

>>>>

Main challenges, motivations
\begin{itemize}
\item In high-order FEM, the curse of dimensionality affects operator
   evaluation and relaxation methods in multigrid and domain-decomposition.  
\item For tensor product cells,
   a tensor product basis will expose the structure to decouple the
   multidimensional problems into a sequence of one-dimensional problems.
   Sum-factorization enables fast operator evaluation, and the fast
   diagonalization method addresses the efficient solution of problems
   structured subdomains.  On unstructured subdomains, an optimal strategy
   using static condensation with FDM bases was proposed in our previous
   work.
\item For simplices, Bernstein--Bezier expansions enable sum-factorized
   operator evaluation, but there is no knowledge of a basis transformation
   that can diagonalize a problem on any structured or unstructured subdomain.
\end{itemize}

@@

For each finite element subcomplex built on a mesh $\mesh$ of a domain $\Omega \subset \mathbb{R}^d$, we denote the spaces discretizing $k$-forms $H\Lambda^k(\Omega)$ with polynomial degree $p$ as $X^k_p(\mesh)$.
These are built from cellwise spaces via the standard pullback construction from a reference cell $\Khat$:
\begin{equation}
X^k_{h,p} = X^k_p(\mesh) = \{ v \in H\Lambda^k(\Omega): \Forall K \in \mesh \Exists \hat{v} \in X^k_p(\Khat) \text{ s.t.} \left.v\right|_K = \pullback^k_K(\hat{v}) \},
\end{equation}
where $\pullback^k_K$ denotes the appropriate pullback that preserves continuity of the traces of a $k$-form across cell facets.

@@

Recall the definition of the $\Hcurl$-conforming finite element spaces
\begin{equation}
   X^1_{p, p_E}(K) = \{v\in [\P_p(K)]^d : \tr v \in \P_{p_E}(E) \Forall E \in \text{edges}(K)  \}
\end{equation}
for $p_E=p-1$ we obtain the \Nedelec space of the first kind $\Ned_p(K)$, while
$p_E=p$ gives the \Nedelec space of the second kind $\NedTwo_p(K) = [\P_p(K)]^d$. 
These spaces enforce continuity of the tangential trace across mesh interfaces.

We define a basis for the dual of $X^1_{p, p_E}(K)$ as 
tangential moments along $E\in \text{edges}(\Khat)$,
\begin{equation}
   \ell^E_j(v) = (q_j, v\cdot \bt)_E, \quad q_j \in \P_{p_E}(E), 
\end{equation}
and for each sub-entity $S \in \text{faces}(\Khat) \cup \text{interior}(\Khat)$,
\begin{align}
   \ell^{S,0}_j(v) &= (\grad_S\phi^S_j, v)_{S}, \\
   \ell^{S,1}_j(v) &= (\curl_S\Phi^S_j, \curl_S v)_{S}, \label{eq:hcurl-interior-dofs}
\end{align}
where $\curl_S$ is the tangential curl on $S$,
$\{\phi^S_j\}$ are the $B^0_{p_S}(S)$ bases constructed in \eqref{eq:hgrad-fdm},
and $\{\curl_S\Phi^S_j\}$ is a basis for $\curl_S B^1_{p_S}(S)$
such that
\begin{equation} \label{eq:hcurl-fdm}
   (\curl_S\Phi^S_j, \curl_S\Phi^S_i)_{S} = \delta_{ij}, \quad
   (\Phi^S_j, \Phi^S_i)_{S} = \lambda_j\delta_{ij}.
\end{equation}

@@

Recall the definition of the $\Hdiv$-conforming finite element spaces
\begin{equation}
   X^2_{p, p_F}(K) = \{v\in [\P_p(K)]^d : \tr v \in \P_{p_F}(F) \Forall F \in \text{faces}(K)  \}
\end{equation}
for $p_F=p-1$ we obtain the Raviart--Thomas space $\RT_p(K)$, while
$p_F=p$ gives the Brezzi--Douglas--Marini space $\BDM_p(K) = [\P_p(K)]^d$. 
These spaces enforce continuity of the normal trace across mesh interfaces.

We define a basis for the dual of $X^2_{p, p_F}(K)$ as normal moments on $F\in
\text{faces}(\Khat)$,
\begin{equation}
   \ell^F_j(v) = (q_j, v\cdot \bn)_F, \quad q_j \in \P_{p_F}(F),
\end{equation}
and on the $\text{interior}(\Khat) = K$: 
\begin{align}
   \ell^{K,0}_j(v) &= (\curl\Phi^K_j, v)_{K}, \\
   \ell^{K,1}_j(v) &= (\div\Psi^K_j, \div v)_{K}, \label{eq:hdiv-interior-dofs}
\end{align}
where $\{\Phi^K_j\}$ are the $B^1_p(K)$ bases constructed in
\eqref{eq:hcurl-fdm}, and $\{\div\Psi^K_j\}$ is a basis for $\div B^2_p(K)$ such that
\begin{equation} \label{eq:hdiv-fdm}
   (\div\Psi^K_j, \div\Psi^K_i)_K = \delta_{ij}, \quad
   (\Psi^K_j, \Psi^K_i)_{K} = \lambda_j\delta_{ij}.
\end{equation}

@@

For 0-forms we construct patches on vertex-stars,
\begin{equation}
   X^0_{h,p} = \left.X^0_{h,p}\right|_\mathcal{I} + X^0_{h,1} 
   + \sum_{v\in E^0(\mesh)} \left.\tilde{X}^0_{h,p}\right|_{\star v} 
\end{equation}

For 1-forms we construct edge-star patches
plus the gradient of vertex-star patches on 0-forms
\begin{equation}
   X^1_{h,p} = \left.X^1_{h,p}\right|_\mathcal{I} +  X^1_{h,1}
   + \sum_{e\in E^1(\mesh)} \left.\tilde{X}^1_{h,p}\right|_{\star e} 
   + \sum_{v\in E^0(\mesh)} \mathrm{d} \left.\tilde{X}^0_{h,p}\right|_{\star v} 
\end{equation}

For 2-forms we construct edge-star patches,
\begin{equation}
   X^2_{h,p} = \left.X^2_{h,p}\right|_\mathcal{I} +  X^2_{h,1} 
   + \sum_{e\in E^1(\mesh)} \left.\tilde{X}^2_{h,p}\right|_{\star e}. 
\end{equation}
